Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-06-16 man/ecosystems/ms_ecosystems.Rmd
update biblio and figures

---
title: "Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges"
output:
  word_document: 
    fig_caption: yes
    fig_height: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
---

@@

* Global change challenges ecosystems in xeric locations transformed by intensive human use. Resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd. populations in the southern Iberian Peninsula was analyzed in relation to historical records of land-use, combining dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.

@@

### Keywords 
extreme drought, resilience, rear edge, *Quercus pyrenaica*, tree growth, dendrochronology, remote sensing

@@

The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016]. It has been often assumed that geographically marginal populations represent ecological marginal populations. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pironon2016; @VilaCabrera2019; @Oldfatheretal2020RangeEdges]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014]. This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and a genetic component [@VilaCabrera2019], but also an anthropogenic (*i.e.* land-use) dimension. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to identify and understand mismatches between the geographical and the ecological rear-edges [@VilaCabreraJump2019GreaterGrowth].

Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sexton2009]. Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allen2010]. Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge. The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@Hampe2005; @VilaCabrera2019].

It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Sala2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @Navarro2013; @DoblasMiranda2017]. In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects]. The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLandUse]. For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests]. Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica]. In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forest at the southern geographical range. Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLandUse], and particularly that of marginal rear edge populations [@VilaCabrera2019].

The assessment of resilience to climate and human disturbances provides critical information concerning the capacity of forests to maintain their structure and render valuable ecosystem services. Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYoua]. @Lloret2011 proposed an approach, which decomposes resilience to drought in three components: resistance to drought, recovery after drought and resilience. Forest resilience is determined by the capacity to reduce the impact (resistance) and the capacity to recover from the impact of disturbance (recovery) [@IngrischBahn2018ComparableQuantification]. This conceptual approach has been very popular to assess forest resilience recently, because it allows a simple, yet highly efficient assessment of short-term responses of trees to drought. Nevertheless, it is not exempted from some criticism and needs to be applied carefully to avoid over interpretation and potential bias associated at different levels [@Schwarzetal2020QuantifyingGrowth]. In this sense we assessed forest resilience both in the long-term to several recent extreme drought episodes and to climate change (*i.e.* warming) in the long-term (the last few decades) using two different proxies to characterize resilience. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and, generally annual tree-ring widths have been used to study growth changes in response to drought at the individual tree level [@Fritts1976; @Dobbertin2005TreeGrowth]. Remote sensing can be used to analyze the impact of drought on ecosystems and hence at the stand level [*e.g.* @Zhang2013]. The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].

In the present study, we assessed resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdo2013; @GeaIzquierdoetal2017RiskyFuture]. We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests. To do so we conducted a review of historical documents to reconstruct forest history in our study sites, and to determine how anthropogenic drivers have shaped the current forest structure. Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations. Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.

```{r, eval=FALSE, echo = FALSE}
# Retales de la intro 
# Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].

@@

# Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].

# Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; @MartinezVilalta2016].

# Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [@Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973]. 
```

## Materials and methods 
### Tree species and study site 
*Quercus pyrenaica* forests extend throughout south-western France and the Iberian Peninsula, reaching their southern limit in mountain areas of northern Morocco [@Franco1990]. In the Iberian Peninsula, these forests occupy siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid, and hyperhumid ombroclimate. Pyrenean oak is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. As a submediterranean species, it has lower drought tolerance than evergreen Mediterranean taxa [@delRio2007].

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability according to elevation and aspect. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002]. Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG]. They also harbor high levels of intraspecific genetic diversity [@Valbuena2013]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013; @GeaIzquierdoetal2017RiskyFuture].

### Climatic data and drought episodes
The Iberian Peninsula underwent several extreme drought episodes in the last three decades [*e.g.* 1995, 1999, 2005, 2012;  @VicenteSerrano2014]. The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], appearing as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3). A drought event was identified using the SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] (SPEI 12-months scale), following a procedure similar to the one proposed by @Spinoni2015. We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: the *duration* as the number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the absolute SPEI values during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

To explore the relationships between climatic variables and tree-growth variables we used climate data obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures had a 0.25 x 0.25 º resolution for the 1950-2016 period. Grid cells were selected to cover each sampled site. The SPEI 6-months scale index was used to characterize the drought conditions for the period 1961-2014.

### Greenness data to assess ecosystem resilience
Vegetation greenness of *Quercus pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor. EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. MODIS EVI data were compiled for the period 2000 - 2016. We selected the pixels covering the distribution of *Quercus pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Any values affected by clouds, snow, shadows or high content aerosols, were filtered out following recommendations for mountain regions [@Reyes2015].

The mean Annual EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, in order to minimize bias in the evaluation of anomalies and to provide more information concerning their magnitude [@Samanta2012]. For each pixel, an annual EVI value was calculated by averaging EVI valid values. Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$, 
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].

Rather than other vegetation indices such as the NDVI, $EVI_{mean}$ was chosen because it is highly stable under the use of any filter [@Reyes2015] and because it showed highly significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88).

### Field sampling and dendroecological methods to assess individual tree resilience
Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect (around 1400 m); and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1). For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit for the species (CA-Low) and the maximum altitude currently reached by trees (CA-High), respectively, in the site sampled. Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled. Two cores of 5 mm in diameter were taken from each tree at breast height (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].

Tree cores were air dried, glued onto wooden mounts, and sanded. Annual radial growth (ring width, RW) was determined with a measuring device coupled to a stereomicroscope, for an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally verified using COFECHA [@Holmes1983].

The growth trends were analyzed at different time scales. To study the growth response to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using dplR [@Bunn2010]. Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976]. The statistical quality of each chronology was checked via the expressed population signal (EPS). A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.

>>>>

---
title: "Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges"
output:
  word_document: 
    fig_caption: yes
    fig_height: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/refs_phd_dendro.bib
csl: ../../refs/ecosystems.csl
---

@@

* Global change challenges ecosystems in xeric locations transformed by intensive human use. Resilience to drought of relict Mediterranean *Quercus pyrenaica* populations in the southern Iberian Peninsula was analyzed in relation to historical records of land-use, combining dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.

@@

### Keywords 
extreme drought, resilience, rear edge, *Quercus pyrenaica*, tree growth, dendroecology, remote sensing

@@

The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@HampePetit2005ConservingBiodiversity]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @SagarinGaines2002AbundantCentre; @Pirononetal2017GeographicVariation]. It has been often assumed that geographically marginal populations represent ecological marginal populations. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pirononetal2017GeographicVariation; @VilaCabreraetal2019RefiningPredictions]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sextonetal2009EvolutionEcology; @Abelietal2014EffectsMarginality; @Oldfatheretal2020RangeEdges]. This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and a genetic component [@VilaCabreraetal2019RefiningPredictions], but also an anthropogenic (*i.e.* land-use) dimension. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to identify and understand mismatches between the geographical and the ecological rear-edges [@VilaCabreraJump2019GreaterGrowth].

Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sextonetal2009EvolutionEcology]. Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allenetal2010GlobalOverview]. Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerranoetal2014EvidenceIncreasing; @Staggeetal2017ObservedDrought]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge. The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@HampePetit2005ConservingBiodiversity; @VilaCabreraetal2019RefiningPredictions].

It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Salaetal2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @NavarroGonzalezetal2013WeightLanduse; @DoblasMirandaetal2017ReviewCombination]. In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects]. The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLanduse]. For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests]. Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica]. In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forests at their southern geographical range. Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLanduse], and particularly that of marginal rear edge populations [@VilaCabreraetal2019RefiningPredictions].

The assessment of resilience to climate and human disturbances provides critical information concerning the capacity of forests to maintain their structure and render valuable ecosystem services. Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYou]. @Lloretetal2011ComponentsTree proposed an approach, which decomposes resilience to drought in three components: resistance to drought, recovery after drought and resilience. Forest resilience is determined by the capacity to reduce the impact (resistance) and the capacity to recover from the impact of disturbance (recovery) [@IngrischBahn2018ComparableQuantification]. This conceptual approach has been very popular to assess forest resilience recently, because it allows a simple, yet highly efficient assessment of short-term responses of trees to drought. Nevertheless, it is not exempted from some criticism and needs to be applied carefully to avoid over interpretation and potential bias associated at different levels [@Schwarzetal2020QuantifyingGrowth]. In this sense we assessed forest resilience both in the long-term to several recent extreme drought episodes and to climate change (*i.e.* warming) in the long-term (the last few decades) using two different proxies to characterize resilience. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and, generally annual tree-ring widths have been used to study growth changes in response to drought at the individual tree level [@Fritts1976TreeRings; @Dobbertin2005TreeGrowth]. Remote sensing can be used to analyze the impact of drought on ecosystems and hence at the stand level [*e.g.* @Zhangetal2013MonitoringEstimating]. The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerranoetal2013ResponseVegetation; @Coulthardetal2017TreeGrowth], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness].

In the present study, we assessed resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture]. We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests. To do so we conducted a review of historical documents to reconstruct forest history in our study sites, and to determine how anthropogenic drivers have shaped the current forest structure. Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations. Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.

```{r, eval=FALSE, echo = FALSE}
# Retales de la intro 
# Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardoetal2018DroughtSensitiveness; @Gazoletal2018ForestResilience].

@@

# Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@HampePetit2005ConservingBiodiversity].

# Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allenetal2010GlobalOverview], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerranoetal2013ResponseVegetation; @MartinezVilalta2016].

# Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @DoradoLinanetal2017LargescaleAtmospheric], and some focused on the resilience to drought of rear-edge populations [@Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973]. 
```

## Materials and methods 
### Tree species and study site 
*Quercus pyrenaica* forests extend throughout south-western France and the Iberian Peninsula, reaching their southern limit in mountain areas of northern Morocco [@Franco1990Quercus]. In the Iberian Peninsula, these forests occupy siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid, and hyperhumid ombroclimate. Pyrenean oak is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. As a submediterranean species, it has lower drought tolerance than evergreen Mediterranean taxa [@delRioetal2007BioclimaticAnalysis].

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability according to elevation and aspect. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olaldeetal2002WhiteOaks]. Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@HampePetit2005ConservingBiodiversity]. They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008PhytosociologicalReview]. They also harbor high levels of intraspecific genetic diversity [@ValbuenaCarabanaGil2013GeneticResilience]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedoetal2002DinamicaEvolutiva]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].

### Climatic data and drought episodes
The Iberian Peninsula underwent several extreme drought episodes in the last three decades [*e.g.* 1995, 1999, 2005, 2012;  @VicenteSerranoetal2014EvidenceIncreasing]. The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoaetal2017DroughtTrends], appearing as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3). A drought event was identified using the SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerranoetal2010MultiscalarDrought] (SPEI 12-months scale), following a procedure similar to the one proposed by @Spinonietal2015EuropeanDrought. We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinonietal2018WillDrought; @Pascoaetal2017DroughtTrends]. A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: the *duration* as the number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the absolute SPEI values during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

To explore the relationships between climatic variables and tree-growth variables we used climate data obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylocketal2008EuropeanDaily]. Monthly precipitation and minimum and maximum temperatures had a 0.25 x 0.25 º resolution for the 1950-2016 period. Grid cells were selected to cover each sampled site. The SPEI 6-months scale index was used to characterize the drought conditions for the period 1961-2014.

### Greenness data to assess ecosystem resilience
Vegetation greenness of *Quercus pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor. EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. MODIS EVI data were compiled for the period 2000 - 2016. We selected the pixels covering the distribution of *Quercus pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Any values affected by clouds, snow, shadows or high content aerosols, were filtered out following recommendations for mountain regions [@ReyesDiezetal2015ImplicacionesFiltrado].

The mean Annual EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, in order to minimize bias in the evaluation of anomalies and to provide more information concerning their magnitude [@Samantaetal2012InterpretationVariations]. For each pixel, an annual EVI value was calculated by averaging EVI valid values. Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$, 
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samantaetal2012InterpretationVariations].

Rather than other vegetation indices such as the NDVI, $EVI_{mean}$ was chosen because it is highly stable under the use of any filter [@ReyesDiezetal2015ImplicacionesFiltrado] and because it showed highly significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88).

### Field sampling and dendroecological methods to assess individual tree resilience
Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect (around 1400 m); and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1). For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit for the species (CA-Low) and the maximum altitude currently reached by trees (CA-High), respectively, in the site sampled. Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuqueetal2015DatasetMIGRAME]. At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled. Two cores of 5 mm in diameter were taken from each tree at breast height (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdoCanellas2009AnalysisHolm for more details].

Tree cores were air dried, glued onto wooden mounts, and sanded. Annual radial growth (ring width, RW) was determined with a measuring device coupled to a stereomicroscope, for an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally verified using COFECHA [@Holmes1983ComputerassistedQuality].

The growth trends were analyzed at different time scales. To study the growth response to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976TreeRings]. Tree-ring width series were standardized and detrended using dplR [@Bunn2010StatisticalVisual]. Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976TreeRings]. The statistical quality of each chronology was checked via the expressed population signal (EPS). A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.
